---
title: "Expo Weasel Model 2025"
output: html_notebook
---

# Parameters

## Files

`working_directory [char]`: Location where all files should be stored. All other paths should be partial paths, using this variable to form full paths.

`PA_obs_path``[char]`: Path to presence-absence observations CSV data.

`PA_locs_path``[char]`: Path to presence-absence location CSV data.

`PO_path``[char]`: Path to presence-only CSV data.-

`shp_path``[char]`: Path to a shapefile file boundary this project.

`covar_paths``[vec(char)]`: A vector of paths to custom (non-WorldClim) covariates to be used. These should be raster files such as tiff images.

## Config

`species [vec(char)]`: A vector of species to include in this analysis.

`max_spatial_error [int/float]`: The maximum spatial distance error (in meters) to include. Records with values higher than this will be filtered out.

`date_cutoff [char]`: The earliest date to include in this analysis (formatted as mm/dd/YYYYY). Records with values lower than this will be filtered out.

`worldclim_covars [vec(char)]`: A vector of WorldClim covariates to include. These use the abbreviated names, such as "elev" for elevation.

`worldclim_res [int/float]`: The resolution of WorldClim covariates. This must be one of the options available from WorldClim, which are provided in minutes. This variable must be an integer or float, not a string.

`pred_dims [vec(int/float, int/float)]`: Dimensions (X, Y) of the output prediction map(s).

`seed [int]`: The seed to use for this script, used for reproducibility.

`proj [int]`: The projection code for spatial data. This variable must be an integer, not a string.

`show_plots [bool]`: A Boolean value representing if plots should be created in cells after finishing certain processes such as adding data.

```{r}
#Files
working_directory <- 'C:\\Users\\tanne\\Desktop\\weaselModeling'
PA_obs_path <- 'Data\\PA_obs.csv'
PA_locs_path <- 'Data\\PA_locs.csv'
PO_path <- 'Data\\PO.csv'
shp_path <- 'Data\\OHBuffer\\Oh_1kmBuffer.shp'
covar_paths <- c('Data\\Covariates\\OH_NLCD_Forest_2023.tif')

#Config
species <- c("Long-tailed weasel", "Short-tailed weasel", "Least weasel")
max_spatial_error <- 10000 #In Meters
date_cutoff <- '1/1/1970'
worldclim_covars = c('elev', 'tavg')
worldclim_res = 2.5
pred_dims <- c(400, 400)
seed <- 890 #Generated from numbergenerator.org/randomnumbergenerator/1-1000
proj <- 4326
show_plots = TRUE
```

# Initialization

## Install and Import Packages

```{r warning=FALSE}
#List of Packages Used
packages <- c('dplyr','lubridate','intSDM','sf','terra','INLA')

#Check if Installed, Install if Not -- Import Regardless
for (pkg in packages) {
  if (!(require(pkg, character.only = TRUE))) {
    if (pkg == 'INLA') {
      install.packages(
        'INLA',
        repos = c(
          getOption("repos"),
          INLA = "https://inla.r-inla-download.org/R/stable"
        ), 
        dep = TRUE
      )
    } else {
      install.packages(pkg)
    }
    
    library(pkg, character.only = TRUE)
  }
}

remove(packages, pkg)
```

## Set Seed and Working Directory

```{r}
#Set Seed
set.seed(seed)

#Set Working Directory
setwd(working_directory)

```

# Data Formatting

## Read in Datasets

```{r}
PA_obs <- read.csv(file = PA_obs_path) 
PA_locs <- read.csv(file = PA_locs_path) 
PO <- read.csv(file = PO_path)
```

```{r}
#Fix Column Types
PO$date <- lubridate::mdy(PO$date)
PA_obs$date <- lubridate::mdy(PA_obs$date)
PA_obs$date2 <- lubridate::mdy(PA_obs$date2)
PA_locs$setDate <- lubridate::mdy(PA_locs$setDate)
PA_locs$retrieveDate <- lubridate::mdy(PA_locs$retrieveDate)

#Create Generalized ID
## Accounts for reused names between projects
## Seperate with "::" incase it needs split later
PA_obs$ID <- paste0(PA_obs$dataSource, "::", PA_obs$camera_site, '::', PA_obs$deployment)
PA_locs$ID <- paste0(PA_locs$dataSource, "::", PA_locs$camera_site, '::', PA_locs$deployment)

#Create Year Field
PO$year <- lubridate::year(PO$date)
PA_obs$year <- lubridate::year(PA_obs$date)
```

## Filter

```{r}
#Presence-only Data
PO <- PO |>
  filter(
    !is.na(latitude), # Rem NULL latitude
    !is.na(longitude), # Remo NULL longitude
    species %in% .env$species, # Rem "Unknown" records
    positionalError_meters < max_spatial_error, #Rem Large Spatial Error
    !(is.na(positionalError_meters)), #Rem Unknown Error
    !('S' %in% positionalError_meters), #Rem Unknown Error
    !(dataNotes %in% c('Quality Grade: needs_id','Quality Grade: casual')), #Rem iNat non-Research Grade
    date > mdy(date_cutoff)
  )

#Presence-Absence Data
PA_obs <- PA_obs |>
  filter(
    species %in% .env$species, #Rem "Unknown" records
  ) 
  
PA_locs <- PA_locs |>
  filter(
    locFuzz == 0, #Rem fuzzed locations
    positionalError_meters < max_spatial_error, #Rem Large Spatial Error
    !(is.na(positionalError_meters)), #Rem Unknown Error
  )
```

## Copy Columns for Presence-Absence Data

```{r}
#Copy Locational Data Columns to Presence-Absence Observations
PA_obs <- merge(
  x = PA_obs, 
  y = PA_locs[,c('ID','camType','latitude','longitude')], 
  by = 'ID', 
  all.x = TRUE
)
```

## Create Occupancy Matrix

```{r}
#Convert to Long Format
PA_long <- PA_obs |>
  group_by(ID, date, species) |>
  summarize(count = n()) |>
  ungroup()
```

```{r}
#Create Columns for All Dates
##Get List of All Dates in Range
dates <- as.character(
  seq(
    min(PA_locs$setDate), 
    max(PA_locs$retrieveDate), 
    by = 'days'
  )
)

##Get List of Sites
sites <- unique(PA_locs$ID)

##Initialize DF
PA_occu <- data.frame(matrix(ncol = 8))

##Add Columns and Row Value
colnames(PA_occu) <- c('ID','date','day_num','species','detection','latitude','longitude', 'camType')
```

```{r}
#Create Matrix
for (i in 1:nrow(PA_locs)) {
  for (sp in species) {
    ##Get List of Days Camera was Active
    dep_days <- as.character(
      seq(
        PA_locs[i,'setDate'],
        PA_locs[i,'retrieveDate'],
        by = 'days'
        )
    )
    
    ##Get Days with Observations for ID + Species
    det_days <- filter(
      PA_long, PA_long$ID == PA_locs[i, 'ID'] & PA_long$species == sp)$date
    
    ##Create Boolean Detection Values
    if (length(det_days) > 0) {
      occu_days <- ifelse(dep_days %in% det_days, 1, 0)
    } else {
      occu_days <- rep(0, length(dep_days))
    }

    ##Create DF
    x <- data.frame(
      ID = rep(PA_locs[i, 'ID'][[1]], length(dep_days)),
      date = dep_days,
      day_num = 1:length(dep_days),
      species = rep(sp, length(dep_days)),
      detection = occu_days,
      latitude = rep(PA_locs[i, 'latitude'][[1]], length(dep_days)),
      longitude = rep(PA_locs[i, 'longitude'][[1]], length(dep_days)),
      camType = rep(PA_locs[i, 'camType'][[1]], length(dep_days))
    )
    
    ##Add DF to DF
    PA_occu <- bind_rows(PA_occu, x)
  }
}

#Remove Empty First Row
PA_occu <- PA_occu[-c(1),] 

remove(x, occu_days, dep_days, det_days, i, sp, dates, sites)
```

# *intSDM* Model Workflow

## Start Workflow

```{r}
#Get Ohio Boundary (+2km Buffer)
shp <- st_read(shp_path) |> st_set_crs(proj)
#shp <- st_set_crs(shp, proj)

#Start Workflow
workflow <- startWorkflow(
  Projection = crs(shp),
  Species = species,
  saveOptions = list(projectName = 'OhioWeasels2025'), 
  Save = FALSE # <----------------
)

#Add Boundary
workflow$addArea(Object = shp)

#Set Model Workflow Outputs
workflow$workflowOutput(c('Model','Maps', 'Cross-validation')) #Should this be moved up? -- was with cross-validation #

#Plot
if (show_plots == TRUE) {workflow$plot()}
```

## Add Occurrence Data

```{r}
#Presence-Only Data
##Convert to ST
PO <- st_as_sf(x = PO, coords = c('longitude','latitude'), crs = crs(shp))
PO <- st_set_crs(PO, proj)

##Add to Workflow
workflow$addStructured(
  dataStructured = PO,
  datasetType = 'PO',
  speciesName = 'species',
  #datasetName = 'Presence Only Data',
  generateAbsences = TRUE
)

#Presence-Absence Data
##Convert to ST
PA_occu <- st_as_sf(x = PA_occu, coords = c('longitude','latitude'), crs = crs(shp)) |> st_set_crs(proj)
#PA_occu <- st_set_crs(PA_occu, proj)

##Add to Workflow
workflow$addStructured(
  dataStructured = PA_occu,
  datasetType = 'PA',
  speciesName = 'species',
  responseName = 'detection',
  trialsName ='day_num'
)

#Plot
if (show_plots == TRUE) {workflow$plot(Species = TRUE)}
```

## Add Covariates

```{r}
#WorldClim Covariates
for (var in worldclim_covars) {
  workflow$addCovariates(worldClim = var, res = worldclim_res, Function = scale)
}

#Add Custom Covariates
for (var in covar_paths) {
  workflow$addCovariates(Object = scale(terra::rast(var)), Function = scale)
}

remove(var)


#for (covar in covar_paths) {
#  workflow$addCovariates(Object = scale(terra::rast(covar)), Function = 'mean')
#}
#remove(covar)
#Plot
if (show_plots == TRUE) {workflow$plot(Species = TRUE, Covariates = TRUE)}
```

## Add Mesh

```{r}
#Add Mesh
workflow$addMesh(
  cutoff = 0.2 * 5,
  max.edge = c(0.1, 0.24) * 80,
  offset = c(0.1, 0.4) * 100
)

#Plot
if (show_plots == TRUE) {workflow$plot(Mesh = TRUE, Boundary = TRUE)}
```

## Specify Priors

```{r}
workflow$specifySpatial(
  prior.range = c(30, 0.1),
  prior.sigma = c(1, 0.1)
)

workflow$specifyPriors(
  effectNames = 'Intercept', 
  Mean = 0, 
  Precision = 1
)
```

## Cross Validation

```{r}
#Cross Validation
workflow$crossValidation(
  Method = 'spatialBlock',  
  blockOptions = list(
    k = 4, 
    rows_cols = c(20, 20), 
    plot = TRUE, 
    seed = seed
  ),
  blockCVType = 'Predict' 
)
```

## Model Output

```{r}
#Create Model
Maps <- sdmWorkflow(
  workflow, 
  inlaOptions = list(
    num.threads = 1,
    control.inla = list(
      int.strategy = 'ccd',
      h = 1e-4,
      cmin = 0,
      control.vb=list(
        enable = FALSE
      )
    ),
      safe = TRUE,
      verbose = TRUE,
      inla.mode = 'experimental'
    ),
  predictionDim = c(400, 400),
  ipointsOptions = list(method = 'direct')
)
```

## Show Maps

This cell sometimes doesn't run correctly and will only show the first map. Selecting the three lines starting with "Maps" and re-running should fix this. I could not figure out how to dynamically update this with the `species` variable, so if this changes you will have to manually change this cell to view the maps in R.

```{r}
if (show_plots == TRUE) {
  Maps$`Long-tailed_weasel`$Maps
  Maps$`Short-tailed_weasel`$Maps
  Maps$`Least_weasel`$Maps
}
```

## Write Distribution Map Data to CSV/Raster

The underlying data is points, but these can be converted to a raster. The raster *could* have some data loss, so both the original points and derived raster are saved. CSV data also includes all fields.

```{r}
#Write to CSV
##Create Output Directory
if (!(dir.exists(paste0(working_directory, '\\', 'DataOutput')))) {
  dir.create(paste0(working_directory, '\\', 'DataOutput'))
}

##Export Data by Species
for (sp in species) {
  sp <- gsub(' ', '_', sp)
  write.csv(
    x = Maps[[sp]][['Maps']][['predictions']][['predictions']][['mean']][['layers']][[1]]$layer_data(),
    file = paste0(working_directory, '\\', 'DataOutput', '\\', sp, '_predictData.csv')
  )
}

remove(sp)
```

```{r}
#Write to Raster
##Create Output Directory
if (!(dir.exists(paste0(working_directory, '\\', 'DataOutput')))) {
  dir.create(paste0(working_directory, '\\', 'DataOutput'))
}

##Export Data by Species
for (sp in species) {
  sp <- gsub(' ', '_', sp)
  
  ### Get Predictions Info + Change Class
  x <- Maps[[sp]][['Predictions']][['predictions']]
  class(x) <- c('sf', 'data.frame') #Not sure if this line is needed
  
  ### Create SpatRaster
  preds <- terra::vect(x)
  
  ### Create Intermediatte Raster
  rast_temp <- terra::rast(preds, ncols = pred_dims[1]-2, nrows = pred_dims[2]-2)
  
  ### Rasterize W/ Correct Field
  rast <- rasterize(preds, rast_temp, field = 'mean')
  
  ### Write to tiff
  writeRaster(x = rast, filename = paste0(working_directory, '\\', 'DataOutput', '\\', sp, '_predictData.tif'), overwrite = TRUE)
  
  
}

remove(sp, x, preds, rast, rast_temp)
```
